\documentclass[12pt,a4paper,twoside,openright]{report}
\usepackage[pdfborder={0 0 0}]{hyperref}
\usepackage[margin=25mm]{geometry}
\usepackage{standalone}

% my imports
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amscd}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{mathpartir}
\usepackage{isabelle}
\usepackage{isabellesym}
\usepackage{comment}
\usepackage[outputdir=output]{minted}

% bibliography
\usepackage[backend=bibtex, sorting=none]{biblatex}
\bibliography{dissertation}
\bibliography{proposal/proposal}

% minted configuration
\usepackage[gray]{xcolor}
\setminted{frame=lines}

\raggedbottom
\sloppy
\clubpenalty1000%
\widowpenalty1000%
\renewcommand{\baselinestretch}{1.1}

\author{Michael Rawson}
\date{2016--17}

% helpful macros
\newcommand{\thetitle}{Verified Metatheory and Type Inference for a Name-Carrying Simply-Typed \(\lambda\)-Calculus}
\newcommand{\ident}[1]{\texttt{#1}}
\newcommand{\infertype}{\mathrm{infer}}
\newcommand{\names}{\mathbb{A}}
\newcommand{\unit}{\mathrm{unit}}
\newcommand{\wrap}[1]{\left(#1\right)}
\newcommand{\swap}{\leftrightarrow}
\newcommand{\dollar}{\mathbin{\textrm{\footnotesize{\textbf{\$}}}}}
\newcommand{\image}{\mathbin{\textrm{\footnotesize{\textbf{\{\$\}}}}}}
\newcommand{\fvs}{\mathrm{fvs}}
\newtheorem*{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newenvironment{implementation}
 {\bigskip\setlength{\leftskip}{1cm}\begin{isabellebody}}
 {\end{isabellebody}\setlength{\leftskip}{0cm}\bigskip}

\begin{document}
\pagestyle{empty}
\rightline{\LARGE\textbf{Michael Rawson}}
\vspace*{60mm}
\begin{center}
\Huge
\textbf{\thetitle}\\[5mm]
Computer Science Tripos -- Part II\\[5mm]
Robinson College\\[5mm]
\today
\end{center}

%proforma
\pagestyle{plain}
\chapter*{Proforma}
{\large
\begin{tabular}{p{0.3\textwidth}p{0.7\textwidth}}
Name:              & \bf Michael Rawson\\
College:           & \bf Robinson College\\
Project Title:     & \bf \thetitle\\
Examination:       & \bf Computer Science Tripos -- Part II, July 2017\\
Word Count:        & 14063\footnotemark[1]\\
Project Originator:& Dr. Dominic Mulligan\\
Supervisors:       & Dr. Dominic Mulligan and Dr. Victor Gomes\\
\end{tabular}
}
\footnotetext[1]{As computed by \emph{texcount} --- uh oh!}
\stepcounter{footnote}

\section*{Original Aims of the Project}
I aim to create a mechanised implementation implementation in Isabelle of the simply-typed \(\lambda\)-calculus, together with a verified algorithm for type inference.
Emphasis is placed on the treatment of binders in the calculus and the approach taken with representing \(\alpha\)-equivalence.

\section*{Work Completed}
This project meets all proposed success criteria, and adds two extensions.
I have implemented the calculus, encoded the \(\alpha\)-equivalence equivalence relation, and specified typing rules and a type inference algorithm for the calculus.
A number of correctness proofs accompany this implementation.
Further to this, I implemented extensions, one to add unit and pair values and corresponding types to the implementation, and one to show the confluence property of the calculus.

\section*{Special Difficulties}
None.

\newpage
\section*{Declaration}
I, Michael Rawson of Robinson College, being a candidate for Part II of the Computer Science Tripos, hereby declare that this dissertation and the work described in it are my own work, unaided except as may be specified below, and that the dissertation does not contain material that has already been used to any substantial extent for a comparable purpose.

\bigskip
\leftline{Signed }
\medskip
\leftline{Date }
\tableofcontents

\newpage
\section*{Acknowledgements}
I wish to acknowledge the efforts of my supervisors in guiding me in this project.
Both offered advice when I needed it, and let me get on when I didn't --- an unusual and invaluable skill!

\pagestyle{headings}

\input{chapters/introduction/introduction}
\input{chapters/preparation/preparation}
\input{chapters/implementation/implementation}
\input{chapters/evaluation/evaluation}
\input{chapters/conclusion/conclusion}
\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography
\appendix
\chapter{Project Proposal}
\input{proposal/proposal}
\end{document}
